---
title: "EDA - Energy Consumption and Weather Data"
author: "Manuel Alejandro Matías Astorga"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: false
---

This notebook performs an exploratory data analysis (EDA) on the [hourly energy demand, generation, pricing, and weather dataset](https://www.kaggle.com/datasets/nicholasjhana/energy-consumption-generation-prices-and-weather).

> This dataset contains four years of hourly data on electricity consumption, generation, prices, and weather conditions in Spain. Energy data was retrieved from ENTSOE (a public TSO portal), pricing data from Red Eléctrica Española, and weather data (for the five largest cities in Spain) was originally obtained via the OpenWeather API and made public by the dataset author.

The main goals of this EDA are:

- To understand the structure and quality of the datasets.
- To detect missing or inconsistent values.
- To identify temporal patterns in energy usage.
- To prepare the data for feature engineering and modeling.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)        # Set global options for knitr
options(scipen = 999)  # avoid scientific notation
```

**📦 Load libraries**

```{r libraries}
# Load necessary packages for data wrangling and visualization
library(tidyverse)    # Core data manipulation tools
library(lubridate)    # Date-time parsing and extraction
library(skimr)        # Data summary and structure
library(janitor)      # Clean column names
library(ggplot2)      # Data visualization
library(ggthemes)     # Visualization themes
library(viridis)      # Color palettes
library(ggcorrplot)   # Correlation matrix visualization
```

**📁 Load data**

```{r load_data}
# Read raw CSV files (energy and weather) from 'data/raw' relative to notebook location
energy <- read_csv("../data/raw/energy_dataset.csv")
weather <- read_csv("../data/raw/weather_features.csv")
```

**🧹 Clean column names**

```{r clean-names}
# Transforming column names to snake_case for consistency
energy <- energy %>% clean_names()
weather <- weather %>% clean_names()
```


## 🧹 1. Data Overview, Quality Check & Cleaning

To begin the analysis, we perform a quick structural inspection using `glimpse()` and a detailed statistical summary with `skim()` for both datasets: **energy** and **weather**.

**🔍 Glimpse: Energy Dataset**

```{r structure-energy}
# Display the structure of the energy dataset
glimpse(energy)
```

The *energy* dataset contains 35,064 hourly records and 29 columns, which include:

- Timestamps (`time`)
- Energy generation by source (e.g., `generation_biomass`, `generation_solar`)
- Forecast and actual values for energy consumption (`total_load_forecast`, `total_load_actual`)
- Pricing variables (`price_day_ahead`, `price_actual`)

The data appears to span from *2014-12-31* to *2018-12-31* with 1-hour intervals.

**🔍 Glimpse: Weather Dataset**

```{r structure-weather}
# Display the structure of the weather dataset
glimpse(weather)
```

The weather dataset has 178,396 rows and 17 columns, including:

- Timestamps (`dt_iso`)
- Weather observations for major Spanish cities (`city_name`)
- Variables such as `temp`, `humidity`, `wind_speed`, `pressure`, and `weather_description`

This dataset also has a datetime range matching the energy dataset, which facilitates temporal merging.

**📊 Skim: Summary Statistics**

**⚡ Energy Dataset**

```{r skim-summary-energy}
# Display summary statistics for the energy dataset
skim(energy)
```

- Most columns are numeric (`dbl`), with one datetime column (`POSIXct`) and two logical columns that represent completely missing data (`NA` only).
- Missing data is generally very low (≤ 0.1%) across most variables, except for two:
  - `generation_hydro_pumped_storage_aggregated`
  - `forecast_wind_offshore_eday_ahead`
These two were removed from the dataset due to having **100% missing values**.
- Some generation sources such as **coal-derived gas**, **oil shale**, and **geothermal** are recorded as zeros or missing throughout the entire time series — this might be expected depending on Spain’s energy infrastructure.

**🌦️ Weather Dataset**

```{r skim-summary-weather}
# Display summary statistics for the weather dataset
skim(weather)
```

- Weather data is well-structured with `no missing values` in any column.
- It includes:
  - Numeric weather conditions (`temp`, `pressure`, `wind_speed`, etc.)
  - Categorical weather types (`weather_main`, `weather_description`)
  - Identifiers and icons from the API (`weather_id`, `weather_icon`)

This dataset is well-suited for modeling, with consistent sampling and granularity.

**📌 Conclusion of Initial Inspection**

- Both datasets are **clean and well-structured**.
- We removed only two columns due to full missingness.
- The time coverage and granularity are compatible, making them ideal for feature engineering and merging.
- From this point forward, we can derive temporal features, visualize trends, and assess correlations.

Finally we treat our datasets based on our observations:

- We double-check for missing values,

```{r missing-values}
# Summarize missing values per variable in both datasets to assess data quality
# NA summary - Energy
energy %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  mutate(
    pct_missing = round(100 * n_missing / nrow(energy), 2)
  ) %>%
  arrange(desc(pct_missing))

# NA summary - Weather
weather %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  mutate(
    pct_missing = round(100 * n_missing / nrow(weather), 2)
  ) %>%
  arrange(desc(pct_missing))
```

- remove columns with 100% missingness and

```{r eliminating-columns-with-100-percent-missing}
# Remove columns that contain 100% missing values and provide no useful information
energy <- energy %>%
  select(-any_of(c("generation_hydro_pumped_storage_aggregated",
                   "forecast_wind_offshore_eday_ahead")))
```

- rename time columns to `datetime` for consistency.

```{r rename-columns-datetime}
# Rename time columns to 'datetime' for consistency
energy <- energy %>%
  rename(datetime = time)
weather <- weather %>%
  rename(datetime = dt_iso)
```

and that's it! We are ready to merge the datasets.

## 🔍 2. Merging Datasets

Once we have cleaned the datasets, we can merge them based on the `datetime` column. This will allow us to analyze the relationship between **energy** consumption and **weather** conditions.

```{r merge}
combined <- left_join(energy, weather, by = "datetime")
```

We use `left_join()` to ensure that all records from the energy dataset are retained, even if there are no corresponding weather records.

## 📊 3. Exploratory Plots

In this section, we visualize major variables over time to understand trends, patterns, and potential anomalies. All plots are based on the `combined` dataset, which merges **weather** and **energy** information.

### 📈 3.1 Energy Load: Forecast vs Actual

This plot shows the forecasted vs actual energy load over time. While the general trends match, we can spot some deviations, which may indicate model drift, unexpected events, or forecasting bias.

```{r plot-energy-load}
combined %>%
  select(datetime, total_load_forecast, total_load_actual) %>%
  pivot_longer(-datetime) %>%
  ggplot(aes(x = datetime, y = value, color = name)) +
  geom_line(alpha = 0.6) +
  labs(title = "Energy Load: Forecast vs Actual", y = "Load (MW)", x = "Time") +
  theme_minimal()
```

### 🌤️ 3.2 Weather Trends Over Time

We track the evolution of temperature, humidity, and wind speed. The periodic structure (particularly in temperature) suggests strong seasonality. Humidity and wind speed show more variability, potentially correlating with energy production in certain sources.

```{r plot-weather-trends}
combined %>%
  select(datetime, temp, wind_speed, humidity) %>%
  pivot_longer(-datetime) %>%
  ggplot(aes(x = datetime, y = value, color = name)) +
  geom_line(alpha = 0.5) +
  facet_wrap(~name, scales = "free_y") +
  labs(title = "Weather Variables Over Time", y = "Value", x = "Time") +
  theme_minimal()
```

### 📊 3.3 Price Trends

This chart compares the day-ahead market price with the actual price. Large discrepancies may highlight forecast errors or market volatility. These trends are essential for assessing model accuracy and energy strategy.

```{r plot-prices}
combined %>%
  select(datetime, price_day_ahead, price_actual) %>%
  pivot_longer(-datetime) %>%
  ggplot(aes(x = datetime, y = value, color = name)) +
  geom_line(alpha = 0.6) +
  labs(title = "Electricity Prices: Day Ahead vs Actual", y = "€/MWh", x = "Time") +
  theme_minimal()
```

### 🔎 3.4 Energy Load - Sample Week

To better understand short-term fluctuations, we zoom into the first week of 2015. We observe clear daily cycles, indicating high periodicity and potential for time-series decomposition.

```{r plot-prices-zoom}
combined %>%
  filter(datetime >= as.POSIXct("2015-01-01"),
         datetime <= as.POSIXct("2015-01-07")) %>%
  ggplot(aes(x = datetime, y = total_load_actual)) +
  geom_line() +
  labs(title = "Energy Load - Sample Week", y = "MW", x = "Date") +
  theme_minimal()
```



📈 4. Correlation Analysis

```{r correlation}
combined %>%
  select(where(is.numeric)) %>%   # just numeric columns
  drop_na() %>%
  cor() %>%
  as.data.frame() %>%
  round(2)
```

```{r correlation-plot}
# Visualize correlation matrix
corr_matrix <- combined %>%
  select(where(is.numeric)) %>%
  drop_na() %>%
  cor()

ggcorrplot(corr_matrix, type = "lower", lab = TRUE)
```

📅 Temporal Patterns

```{r time-features}
combined <- combined %>%
  mutate(
    hour = hour(datetime),
    wday = wday(datetime, label = TRUE),
    month = month(datetime, label = TRUE)
  )
```

🧠 5. Insights & Next Steps


