---
title: "EDA - Energy Consumption and Weather Data"
author: "Manuel Alejandro Matías Astorga"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: false
---

This notebook performs an exploratory data analysis (EDA) on the [hourly energy demand, generation, pricing, and weather dataset](https://www.kaggle.com/datasets/nicholasjhana/energy-consumption-generation-prices-and-weather).

> This dataset contains four years of hourly data on electricity consumption, generation, prices, and weather conditions in Spain. Energy data was retrieved from ENTSOE (a public TSO portal), pricing data from Red Eléctrica Española, and weather data (for the five largest cities in Spain) was originally obtained via the OpenWeather API and made public by the dataset author.

The main goals of this EDA are:

- To understand the structure and quality of the datasets.
- To detect missing or inconsistent values.
- To identify temporal patterns in energy usage.
- To prepare the data for feature engineering and modeling.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)        # Set global options for knitr
options(scipen = 999)  # avoid scientific notation
```

**📦 Load libraries**

```{r libraries}
# Load necessary packages for data wrangling and visualization
library(tidyverse)    # Core data manipulation tools
library(lubridate)    # Date-time parsing and extraction
library(skimr)        # Data summary and structure
library(janitor)      # Clean column names
library(ggplot2)      # Data visualization
library(ggthemes)     # Visualization themes
library(viridis)      # Color palettes
library(ggcorrplot)   # Correlation matrix visualization
```

**📁 Load data**

```{r load_data}
# Read raw CSV files (energy and weather) from 'data/raw' relative to notebook location
energy <- read_csv("../data/raw/energy_dataset.csv")
weather <- read_csv("../data/raw/weather_features.csv")
```

## 🧹 1. Data Overview & Structure

To begin the analysis, we perform a quick structural inspection using `glimpse()` and a detailed statistical summary with `skim()` for both datasets: **energy** and **weather**.

**🔍 Glimpse: Energy Dataset**

```{r structure-energy}
# Display the structure of the energy dataset
glimpse(energy)
```

The *energy* dataset contains 35,064 hourly records and 29 columns, which include:

- Timestamps (`time`)
- Energy generation by source (e.g., `generation_biomass`, `generation_solar`)
- Forecast and actual values for energy consumption (`total_load_forecast`, `total_load_actual`)
- Pricing variables (`price_day_ahead`, `price_actual`)

The data appears to span from *2014-12-31* to *2018-12-31* with 1-hour intervals.

**🔍 Glimpse: Weather Dataset**

```{r structure-weather}
glimpse(weather)
```

The weather dataset has 178,396 rows and 17 columns, including:

- Timestamps (`dt_iso`)
- Weather observations for major Spanish cities (`city_name`)
- Variables such as `temp`, `humidity`, `wind_speed`, `pressure`, and `weather_description`

This dataset also has a datetime range matching the energy dataset, which facilitates temporal merging.

**📊 Skim: Summary Statistics**

**⚡ Energy Dataset**

```{r skim-summary-energy}
skim(energy)
```

- Most columns are numeric (`dbl`), with one datetime column (`POSIXct`) and two logical columns that represent completely missing data (`NA` only).
- Missing data is generally very low (≤ 0.1%) across most variables, except for two:
  - `generation_hydro_pumped_storage_aggregated`
  - `forecast_wind_offshore_eday_ahead`
These two were removed from the dataset due to having **100% missing values**.
- Some generation sources such as **coal-derived gas**, **oil shale**, and **geothermal** are recorded as zeros or missing throughout the entire time series — this might be expected depending on Spain’s energy infrastructure.

**🌦️ Weather Dataset**

```{r skim-summary-weather}
skim(weather)
```

- Weather data is well-structured with `no missing values` in any column.
- It includes:
  - Numeric weather conditions (`temp`, `pressure`, `wind_speed`, etc.)
  - Categorical weather types (`weather_main`, `weather_description`)
  - Identifiers and icons from the API (`weather_id`, `weather_icon`)

This dataset is well-suited for modeling, with consistent sampling and granularity.

**📌 Conclusion of Initial Inspection**

- Both datasets are **clean and well-structured**.
- We removed only two columns due to full missingness.
- The time coverage and granularity are compatible, making them ideal for feature engineering and merging.
- From this point forward, we can derive temporal features, visualize trends, and assess correlations.

Finally we treat our datasets based on out observations:

- We double-check for missing values and

```{r missing-values}
# Summarize missing values per variable in both datasets to assess data quality
weather %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  mutate(
    pct_missing = round(100 * n_missing / nrow(weather), 2)
  ) %>%
  arrange(desc(pct_missing))

# NA summary - Energy
energy %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  mutate(
    pct_missing = round(100 * n_missing / nrow(energy), 2)
  ) %>%
  arrange(desc(pct_missing))
```

- remove columns with 100% missingness.

```{r eliminating-columns-with-100-percent-missing}
# Remove columns that contain 100% missing values and provide no useful information
# Column names must be cleaned before attempting to remove NA columns
# 1. Clean column names
energy <- energy %>% clean_names()
weather <- weather %>% clean_names()

# 2. Remove 100% missing columns (now using cleaned names)
energy <- energy %>%
  select(-any_of(c("generation_hydro_pumped_storage_aggregated",
                   "forecast_wind_offshore_eday_ahead")))
```

🔍 2. Cleaning & Preprocessing

```{r columns-names}
colnames(energy)
colnames(weather)
```

```{r rename-columns-datetime}
# Rename time columns to 'datetime' for consistency
energy <- energy %>%
  rename(datetime = time)

weather <- weather %>%
  rename(datetime = dt_iso)
```

📊 3. Exploratory Plots

Energy: total production vs consumption

```{r plot-energy}
# Plot actual vs forecasted energy load over time
energy %>%
  select(datetime, total_load_actual, total_load_forecast) %>%
  pivot_longer(-datetime) %>%
  ggplot(aes(x = datetime, y = value, color = name)) +
  geom_line(alpha = 0.7) +
  labs(title = "Energy Load: Actual vs Forecast", y = "Load (MW)", x = "Time") +
  theme_minimal()
```

Weather variables

```{r plot-weather}
# Compare temperature, humidity, and wind speed trends
weather %>%
  select(datetime, temp, wind_speed, humidity) %>%
  pivot_longer(-datetime) %>%
  ggplot(aes(x = datetime, y = value, color = name)) +
  geom_line(alpha = 0.6) +
  facet_wrap(~name, scales = "free_y") +
  labs(title = "Weather Features Over Time") +
  theme_minimal()
```

🔁 4. Merging Datasets (if needed)

```{r merge}
combined <- left_join(energy, weather, by = "datetime")
```

📈 5. Correlation Analysis

```{r correlation}
combined %>%
  select(where(is.numeric)) %>%   # just numeric columns
  drop_na() %>%
  cor() %>%
  as.data.frame() %>%
  round(2)
```

```{r correlation-plot}
# Visualize correlation matrix
corr_matrix <- combined %>%
  select(where(is.numeric)) %>%
  drop_na() %>%
  cor()

ggcorrplot(corr_matrix, type = "lower", lab = TRUE)
```

📅 6. Temporal Patterns

```{r time-features}
combined <- combined %>%
  mutate(
    hour = hour(datetime),
    wday = wday(datetime, label = TRUE),
    month = month(datetime, label = TRUE)
  )
```

🧠 7. Insights & Next Steps


